Dynamics of Colombo's Top: Tidal Dissipation and Resonance Capture

### 2021-04-26 writeups/Su_weak_tides.tex
made a decent amount of progress on weak tides draft...

% label C_\pm on fig 2
% unify 3 & 4, choose consistent q/cos q

% S2: CS stability
% S3: How to reach CS
% S3: toy problem, affords analytical understanding
% S4: full problem
% Discussion/conclusion: maybe 0.05 AU + SE? does this still work?
    \documentclass[
        fleqn,
        usenatbib,
        % referee,
    ]{mnras}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        ae, aecompl,
        graphicx,
        booktabs,
        xcolor,
    }

@@

The equilibria of Eq.~\eqref{eq:dsdt_rot} are referred to as \emph{Cassini
States} \citep[CSs;][]{colombo1966, peale1969}. We follow the notation of
\citet{su2020}, where the parameter
\begin{equation}
    \eta \equiv -\frac{g}{\alpha},\label{eq:def_eta}
\end{equation}
is used. For a given value of $\eta$, there can be either two or four CSs, all
of which require $\uv{s}$ lie in the plane of $\uv{l}$ and $\uv{l}_{\rm p}$.
Following the standard convention and nomenclature, CSs 1, 3, and 4 have $\phi =
0$ and $\theta < 0$, while CS2 has $\phi = \pi$ and $\theta > 0$.
Figure~\ref{fig:cs_locs} shows the CS obliquities as a function of $\eta$, each
of which satisfies
\begin{equation}
    \sin \theta \cos \theta - \eta \sin\p{\theta - I} = 0.\label{eq:cs_zero}
\end{equation}
Note that there are four CSs when $\eta < \eta_{\rm c}$ and only two when $\eta
> \eta_{\rm c}$, where
\begin{equation}
    \eta_{\rm c} \equiv \p{\sin^{2/3}\!I + \cos^{2/3}\!I}^{-3/2}.\label{eq:etac}
\end{equation}
\begin{figure}
    \centering
    %\includegraphics[width=\columnwidth]{../initial/99_misc/2_cs_locs.png}
    \caption{Cassini State obliquities $\theta$ as a function of $\eta \equiv
    -g/\alpha$ (Eq.~\ref{eq:def_eta}) for $I = 5^\circ$. The vertical dashed
    line denotes $\eta_{\rm c}$, where the number of Cassini States changes from
    four to just two (Eq.~\ref{eq:etac}). The horizontal dashed line gives
    $\theta = I$, the asymptotic value of Cassini State 2's obliquity when $\eta
    \to \infty$.}\label{fig:cs_locs}
\end{figure}

The Hamiltonian corresponding to Eq.~\eqref{eq:dsdt_rot} is
\begin{align}
    H &= -\frac{\alpha}{2}\p{\uv{s} \cdot \uv{l}}^2
            - g\p{\uv{s} \cdot \uv{l}_{\rm d}}\nonumber\\
        &= -\frac{\alpha}{2} \cos^2\theta
            - g\p{\cos\theta \cos I - \sin I \sin\theta \cos \phi}.\label{eq:H}
\end{align}
Here, $\cos \theta$ and $\phi$ form a canonically conjugate pair of variables.
Figure~\ref{fig:1contours} shows the level curves of this Hamiltonian for $I =
20^\circ$, for which $\eta_{\rm c} \approx 0.77$ (Eq.~\ref{eq:etac}). When $\eta
< \eta_{\rm c}$, CS4 exists and is a saddle point. The two trajectories
originating and ending at CS4 are the only two infinite-period orbits in the
phase space. Together, these two critical trajectories are referred to as the
\emph{separatrix} and divide phase space into three zones. Trajectories in zone
II librate about CS2 while those in zones I and III circulate. On the other
hand, when $\eta > \eta_{\rm c}$, all trajectories circulate. When the
separatrix exists, we divide it into two curves: $\mathcal{C}_+$, the boundary
between zones I and II, and $\mathcal{C}_-$, the boundary between zones II and
III\@.
\begin{figure}
    \centering
    %\includegraphics[width=\columnwidth]{../initial/0_eta/1contours.png}
    \caption{Level curves of the Cassini State Hamiltonian (Eq.~\ref{eq:H}) for
    $I = 5^\circ$, for which $\eta_{\rm c} \approx 0.77$ (Eq.~\ref{eq:etac}).
    For $\eta < \eta_{\rm c}$, there are four Cassini States (labeled), while
    for $\eta > \eta_{\rm c}$ there are only two. In the former case, the
    existence of a \emph{separatrix} (solid black lines) separates phase space
    into three numbered zones (I/II/III, labeled). Finally, we denote the upper
    and lower legs of the separatrix by $\mathcal{C}_{\pm}$ respectively, as
    shown in the upper two panels.}\label{fig:1contours}
\end{figure}

\section{Spin Evolution with Alignment Torque}\label{ss:toy_model}

In this section, we consider a simplified model of equilibrium tides that
isolates the important new phenomenon presented in this paper. We assume that
the spin magnitude of the planet is constant, so $\alpha$ and $g$ are both
fixed, while the spin orientation $\uv{s}$ experiences a torque that drives it
towards alignment with $\uv{l}$ on the alignment timescale $t_{\rm s}$:
\begin{equation}
    \p{\rd{\uv{s}}{t}}_{\rm tide}
        = \frac{1}{t_{\rm s}} \uv{s} \times \p{\uv{l} \times \uv{s}}.
\end{equation}
The full equations of motion for $\uv{s}$ in the coordinates $\theta$ and $\phi$
can be written:
\begin{align}
    \rd{\theta}{t} &= -g\sin I \sin \phi - \frac{1}{t_{\rm s}} \sin \theta,
        \label{eq:dqdt_toy}\\
    \rd{\phi}{t} &= -\alpha \cos\theta
        - g\p{\cos I + \sin I \cot \theta \cos \phi}.\label{eq:dfdt_toy}
\end{align}

\subsection{Shifted Cassini States and Linear Stability
Analysis}\label{ss:tidal_eqs}

When $t_{\rm s}$ is finite, the CSs tilt out of the $\uv{l}$-$\uv{l}_{\rm p}$
plane \citep[as was first pointed out for CS2 by][]{fabrycky_otides}. A simple
expression for the degree of this tilt can be obtained by solving for $\phi_{\rm
cs}$, the azimuthal angle of the modified CS, when setting Eq.~\eqref{eq:dqdt_toy} equal
to zero. This gives
\begin{equation}
    \sin \phi_{\rm cs} = \frac{\sin^2\theta_{\rm cs}}{\sin I \abs{g}t_{\rm s}}.
\end{equation}
Here, $\theta_{\rm cs}$ is the modified CS obliquity. Note that if $t_{\rm s}$
is shorter than some critical value $t_{\rm s, c}$, which for a given CS is
given by
\begin{align}
    t_{\rm s, c} &= \frac{\sin^2\theta_{\rm cs}}{\abs{g}\sin I},
\end{align}
then there are no solutions for $\phi_{\rm cs}$, and the CS is no longer a fixed
point. This condition first breaks down for $\theta_{\rm cs} \approx 90^\circ$,
i.e.\ for CS2 (and CS4) in the limit $\eta \ll 1$. Thus, as long as $t_{\rm s}
\gtrsim \p{\abs{g}\sin I}^{-1}$, then all CSs will exist.

We next seek to characterize the stability of small perturbations about each of
the CSs in the presence of the alignment torque. We can linearize
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}) about a [modified] CS, yelding
\begin{align}
    \rd{}{\tau}\begin{bmatrix}
        \theta\\ \phi
    \end{bmatrix} &= \begin{bmatrix}
        -\frac{\cos \theta}{t_{\rm s}} &
        -g\sin I \cos \phi \\
        \alpha \sin \theta + g\frac{\sin I \cos \phi}{\sin^2\theta} &
        0
    \end{bmatrix}_{\rm cs}\begin{bmatrix}
        \Delta \theta \\ \Delta \phi
    \end{bmatrix},\label{eq:dsdt_hessian}
\end{align}
where the cs subscript indicates evaluating at a CS, $\Delta \theta = \theta -
\theta_{\rm cs}$, and the same for $\Delta \phi$. The eigenvalues $\lambda$ of
Eq.~\eqref{eq:dsdt_hessian} satisfy the equation
\begin{align}
    0 &= \p{\lambda + \frac{\cos \theta_{\rm cs}}{t_{\rm s}}}\lambda - \lambda_0^2,\\
    \lambda_0^2 &\equiv \p{\alpha
        \sin \theta_{\rm cs} + g\sin I \csc^2\theta_{\rm cs}\cos \phi_{\rm cs}}
            \p{- g \sin I \cos \phi_{\rm cs}},\\
    \lambda &\approx -\frac{\cos \theta_{\rm cs}}{t_{\rm s}} \pm \sqrt{\lambda_0^2}.
        \label{eq:lambda_approx}
\end{align}
The stability depends only on the real part of
$\lambda$ in Eq.~\eqref{eq:lambda_approx}. If we assume that the torque-induced
modification to $\phi_{\rm cs}$ is small, then $\lambda_0$ reduces to the
product of the eigenvalues in the torque-free limit. In this limit, $\lambda_0^2
< 0$ for each of CSs 1--3 and $\lambda_0^2 > 0$ for CS4 \citep[see Fig.~17
of][]{su2020}. Thus, CS4 is always unstable, since there will always be a
positive eigenvalue $\lambda$, while the stability of CSs 1--3 are solely
determined by the sign of $\cos \theta_{\rm cs}$. From Fig.~\ref{fig:cs_locs},
we see that CSs 1 and 2 are stable while CS3 is unstable under the alignment
torque. These calculations justify results long used in CS literature
\citep[e.g.][]{ward1975tidal}.

\subsection{Spin Obliquity Evolution Driven by Alignment Torque}

With the above results, we are equipped to ask the key question of this section:
for a given initial $\theta_0$ and $\phi_0$, which of CS1 and CS2 does the
system evolve into? For initial conditions in Zone I (see
Fig.~\ref{fig:1contours}), an unperturbed trajectory circulates and returns to
$(\theta_0, \phi_0 + 2\pi)$ after a single precession period. During this
period, the effect of the dissipation is a negative $\theta'$ everywhere during
the cycle. Thus, for initial conditions in Zone I, $\theta$ decreases until the
trajectory has converged to CS1. This is intuitively reasonable, as CS1 is
stable. For initial conditions in Zone II, an equivalent calculation to that in
Section~\ref{sss:cs_stab2} shows that all trajectories converge to CS2.

For illustrative purposes, we fix $\abs{g}t_{\rm s} = 10^{-3}$, and $\eta =
-g / \alpha$ is varied among a few small values (small
meaning much less than $\eta_{\rm c}$). We specialize to the case where $\eta
< \eta_{\rm c}$ and there are four CSs.

However, initial conditions in Zone III pose a puzzle. There are no stable CSs
in Zone III, so the system must evolve through the separatrix to reach one of
either CS1 or CS2. In Fig.~\ref{fig:toy_phop}, we find see that the outcome for
initial conditions in Zone III is in fact \emph{probabilistic}. Intuitively,
this can be understood as probabilistic resonance capture: since $\eta \ll
\eta_{\rm c}$, $\alpha \gg -g$ (the spin-orbit precession rate,
Eq.~\ref{eq:dsdt1}, and the orbit precession induced by the perturber,
Eq.~\ref{eq:dldt1}, respectively), but $\alpha \cos \theta$ can become
commensurate with $-g$ if $\cos \theta \sim \eta$. This is achieved as $\theta$
evolves from an initially retrograde obliquity through $90^\circ$ towards
$0^\circ$ under the influence of the dissipative term in
Eq.~\eqref{eq:dqdt_toy}.
\begin{figure}
    \centering
    %\includegraphics[width=\columnwidth]{../initial/0_eta/3stats3_5_0_2.png}
    \caption{TODO generate a better plot. Ignore bottom two panels, top two
    panels show which ICs end up in CS 1 (left) and CS2 (right). Probabilistic
    behavior in Zone III is evident.}\label{fig:toy_phop}
\end{figure}

While similar in behavior to previous studies of probabilistic resonance capture
\citep{henrard1982, su2020}, the mechanism at play here is necessarily
different: here, the Hamiltonian and underlying phase structure is kept constant
while a non-Hamiltonian, dissipative perturbation is introduced. The origin of
this resonance crossing process is easy to visualize. In the absence of
dissipative effects, the separatrix is both the set of points that flow into CS4
and the set of points that flow away CS4, called the \emph{stable} and
\emph{unstable} manifolds of the saddle point CS4 respectively \citep{g_and_h}.
When the dissipative term is taken into account, these stable and unstable
manifolds split and open a visible ``path'' from Zone III into both Zones I and
II, as shown in Fig.~\ref{fig:toy_hop_manifolds}.
\begin{figure}
    \centering
    %\includegraphics[width=\columnwidth]{../initial/0_eta/6manifolds0_05.png}
    \caption{TODO overplot separatrix, remove bad labels, better legend. Stable
    and unstable manifolds of CS4, showing how the separatrix splrits and gives
    a ``path'' from Zone III into Zones I/II.}\label{fig:}
\end{figure}

The amount that the stable and unstable manifolds split can be calculated
perturbatively using \emph{Melnikov's Method} \citep{g_and_h}. The essence of
Melnikov's method lies at the heart of the perturbative approach used in
Section~\ref{sss:cs_stab2}, where the evolution of $H$ along the perturbed
trajectories can be evaluated by integration along their original
\emph{unperturbed} trajectories. In particular: \textcolor{red}{TODO the rest of
this section is copy pasted from notes without even notation changes, needs to
be written up correctly} Now, let's reconsider the Melnikov integral when
perturbing this finite (non-homoclinic orbit); this may no longer be an exact
result but should give the correct scaling:
\begin{equation}
    M_c = \int\limits_0^T \pd{\phi}{t}\epsilon \p{1 - \mu^2}\;\mathrm{d}t.
\end{equation}
Naively, we might claim that, since $\pd{\phi}{t}$ changes signs halfway through
the interval of integration, that the only surviving component is $2\epsilon
\bar{\mu}\mu'$, where $\bar{\mu} = \mu_4$ is the average value of $\mu$ and
$\mu'$ is the fluctuation. This gives
\begin{equation}
    M_c = 2\int\limits_0^{2\pi}2\epsilon \frac{\eta \cos I}{1 + \eta \sin I}
        \sqrt{2\eta \sin I \p{1 - \cos \phi}}\;\mathrm{d}\phi.
\end{equation}
Note that $M_c \propto \eta^{3/2}$, and since the gap opened $\Delta_c =
\frac{M_c}{\pd{\phi}{t}} \propto \eta$, it seems like we're on the right track.
Specifically:
\begin{align}
    M_c &\approx \epsilon 2\eta \cos I A_{sep},\\
    \Delta_c &\approx 2\epsilon \eta \cos I \p{16 \sqrt{\eta \sin I}}
        \frac{1}{\sqrt{4\eta \sin I}},\\
        &\approx 16\epsilon \eta \cos I.
\end{align}
This also agrees exceedingly well with our simulations This then gives us
hopping probability
\begin{equation}
    P_{hop} = \frac{\Delta_c}{\Delta_M} \approx
        \frac{16\eta^{3/2}\cos I \sqrt{\sin I}}{\pi}.
\end{equation}
This agrees perfectly with the cases we've run.

\textbf{Note:} The full formula, by actually evaluating all the terms, comes out
to be
\begin{equation}
    P_{hop} = \frac{16\eta^{3/2}\cos I \sqrt{\sin I}}{\pi
        \p{1 - 2\eta \sin I - \eta^2 \cos^2 I}
        + 8\eta^{3/2}\cos I \sqrt{\sin I}}.
\end{equation}

\section{Spin Evolution with Weak Tidal Friction}\label{ss:full_tide_prob}

\begin{figure}
    \centering
    %\includegraphics[width=\columnwidth]{../initial/1_weaktide/6equils0_06.png}
    \caption{Schematic illustrating the evolution of the planet's spin due to
    tidal friction. The green lines illustrate the directions of
    Eqs.~(\ref{eq:dsdt_tide}--\ref{eq:dWsdt_tide}), while the black and blue
    lines denote where the tidal $\Omega_{\rm s}'$ and $\theta'$ change signs
    respectively. Overlaid are the obliquities of Cassini States 1 and 2, the
    two Cassini States that are stable under tidal dissipation (see
    Section~\ref{ss:tidal_eqs}). The points that are both Cassini States and
    satisfy $\Omega_{\rm s}' = 0$ are the tidal Cassini Equilibria (tCE),
    circled in green. Finally, the vertical red lines illustrate the
    $\Omega_{\rm s}$ below which tides causes tCE2 to become unstable; the
    dashed and dotted lines correspond to values of $\epsilon = 10^{-2}$ and
    $10^{-1}$ respectively.
    }\label{fig:6equils}
\end{figure}

\subsection{Tidal Model: Equilibrium Tides}\label{sss:weaktide}

>>>>

    \documentclass[
        fleqn,
        usenatbib,
        % referee,
    ]{mnras}
    \usepackage{
        amsmath,
        amssymb,
        newtxtext,
        newtxmath,
        ae, aecompl,
        graphicx,
        booktabs,
        xcolor,
    }

@@

The equilibria of Eq.~\eqref{eq:dsdt_rot} are referred to as \emph{Cassini
States} \citep[CSs;][]{colombo1966, peale1969}. We follow the notation of
\citet{su2020}, where the parameter
\begin{equation}
    \eta \equiv -\frac{g}{\alpha},\label{eq:def_eta}
\end{equation}
is used. For a given value of $\eta$, there can be either two or four CSs, all
of which require $\uv{s}$ lie in the plane of $\uv{l}$ and $\uv{l}_{\rm p}$.
Following the standard convention and nomenclature, CSs 1, 3, and 4 have $\phi =
0$ and $\theta < 0$, while CS2 has $\phi = \pi$ and $\theta > 0$.
Figure~\ref{fig:cs_locs} shows the CS obliquities as a function of $\eta$, each
of which satisfies
\begin{equation}
    \sin \theta \cos \theta - \eta \sin\p{\theta - I} = 0.\label{eq:cs_zero}
\end{equation}
Note that there are four CSs when $\eta < \eta_{\rm c}$ and only two when $\eta
> \eta_{\rm c}$, where
\begin{equation}
    \eta_{\rm c} \equiv \p{\sin^{2/3}\!I + \cos^{2/3}\!I}^{-3/2}.\label{eq:def_etac}
\end{equation}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/2_cs_locs.png}
    \caption{Cassini State obliquities $\theta$ as a function of $\eta \equiv
    -g/\alpha$ (Eq.~\ref{eq:def_eta}) for $I = 5^\circ$. The vertical dashed
    line denotes $\eta_{\rm c}$, where the number of Cassini States changes from
    four to just two (Eq.~\ref{eq:def_etac}). The horizontal dashed line gives
    $\theta = I$, the asymptotic value of Cassini State 2's obliquity when $\eta
    \to \infty$.}\label{fig:cs_locs}
\end{figure}

The Hamiltonian corresponding to Eq.~\eqref{eq:dsdt_rot} is
\begin{align}
    H &= -\frac{\alpha}{2}\p{\uv{s} \cdot \uv{l}}^2
            - g\p{\uv{s} \cdot \uv{l}_{\rm d}}\nonumber\\
        &= -\frac{\alpha}{2} \cos^2\theta
            - g\p{\cos\theta \cos I - \sin I \sin\theta \cos \phi}.\label{eq:H}
\end{align}
Here, $\cos \theta$ and $\phi$ form a canonically conjugate pair of variables.
Figure~\ref{fig:1contours} shows the level curves of this Hamiltonian for $I =
20^\circ$, for which $\eta_{\rm c} \approx 0.77$ (Eq.~\ref{eq:def_etac}). When $\eta
< \eta_{\rm c}$, CS4 exists and is a saddle point. The two trajectories
originating and ending at CS4 are the only two infinite-period orbits in the
phase space. Together, these two critical trajectories are referred to as the
\emph{separatrix} and divide phase space into three zones. Trajectories in zone
II librate about CS2 while those in zones I and III circulate. On the other
hand, when $\eta > \eta_{\rm c}$, all trajectories circulate. When the
separatrix exists, we divide it into two curves: $\mathcal{C}_+$, the boundary
between zones I and II, and $\mathcal{C}_-$, the boundary between zones II and
III\@.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/0_eta/1contours.png}
    \caption{Level curves of the Cassini State Hamiltonian (Eq.~\ref{eq:H}) for
    $I = 5^\circ$, for which $\eta_{\rm c} \approx 0.77$ (Eq.~\ref{eq:def_etac}).
    For $\eta < \eta_{\rm c}$, there are four Cassini States (labeled), while
    for $\eta > \eta_{\rm c}$ there are only two. In the former case, the
    existence of a \emph{separatrix} (solid black lines) separates phase space
    into three numbered zones (I/II/III, labeled). Finally, we denote the upper
    and lower legs of the separatrix by $\mathcal{C}_{\pm}$ respectively, as
    shown in the upper two panels.}\label{fig:1contours}
\end{figure}

\section{Spin Evolution with Alignment Torque}\label{s:toy_model}

In this section, we consider a simplified model of equilibrium tides that
isolates the important new phenomenon presented in this paper. We assume that
the spin magnitude of the planet is constant, so $\alpha$ and $g$ are both
fixed, while the spin orientation $\uv{s}$ experiences an alignment torque
towards $\uv{l}$ on the alignment timescale $t_{\rm s}$:
\begin{equation}
    \p{\rd{\uv{s}}{t}}_{\rm tide}
        = \frac{1}{t_{\rm s}} \uv{s} \times \p{\uv{l} \times \uv{s}}.
        \label{eq:dsdt_tide_toy}
\end{equation}
The full equations of motion for $\uv{s}$ in the coordinates $\theta$ and $\phi$
can be written:
\begin{align}
    \rd{\theta}{t} &= -g\sin I \sin \phi - \frac{1}{t_{\rm s}} \sin \theta,
        \label{eq:dqdt_toy}\\
    \rd{\phi}{t} &= -\alpha \cos\theta
        - g\p{\cos I + \sin I \cot \theta \cos \phi}.\label{eq:dfdt_toy}
\end{align}

\subsection{Shifted Cassini States and Linear Stability
Analysis}\label{ss:tidal_equils}

If the alignment torque is weak ($\abs{gt_{\rm s}} \gg 1$), then the fixed points of
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}) are just slightly shifted CSs. This
shift can be calculated cleanly to leading order: all of the CS obliquities
$\theta_{\rm cs}$ are unchanged while the azimuthal angle $\phi_{\rm cs}$ for
each CS now satisfies
\begin{equation}
    \sin \phi_{\rm cs} = \frac{\sin\theta_{\rm cs}}{\sin I \abs{g}t_{\rm s}}.
        \label{eq:mcs_shift}
\end{equation}
We can see that if $t_{\rm s} > t_{\rm s, c}$, where
\begin{align}
    t_{\rm s, c} &= \frac{1}{\abs{g}\sin I},
\end{align}
then Eq.~\eqref{eq:mcs_shift} will always have solutions for $\phi_{\rm cs}$,
and the alignment torque never changes the number of fixed points of the system.
If $t_{\rm s}$ is decreased below $t_{\rm s, c}$, CS2 and CS4 cease to be
fixed points if $\eta \ll 1$ \citep[as first noted in][]{fabrycky_otides}, as
$\theta_{\rm cs} \approx 90^\circ$ for these (see Fig.~\ref{fig:cs_locs}), while
the other CSs have small $\sin \theta_{\rm cs}$ and are only slightly perturbed.
Figure~\ref{fig:mcs} shows the obliquity and azimuthal angles for each of the
CSs in the $\eta \ll 1$ case, where it can be seen that CS2 and CS4 collide and
annihilate.. For the remainder of this section, we will consider the case where
$t_{\rm s} \gg t_{\rm s, c}$ and the CSs only differ slightly from their
unperturbed locations.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/0_stab.png}
    \caption{Modified CSs for $I = 5^\circ$ and $\eta = 0.2$, and the CS1 and
    CS3 obliquities have been offset (labeled in legend) to improve clarity of
    the plot.}\label{fig:mcs}
\end{figure}

We next seek to characterize the stability of small perturbations about each of
the CSs in the presence of the weak tidal alignment torque. We can linearize
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}) about a shifted CS, yelding
\begin{align}
    \rd{}{t}\begin{bmatrix}
        \theta\\ \phi
    \end{bmatrix} &= \begin{bmatrix}
        -\frac{\cos \theta}{t_{\rm s}} &
        -g\sin I \cos \phi \\
        \alpha \sin \theta + g\frac{\sin I \cos \phi}{\sin^2\theta} &
        0
    \end{bmatrix}_{\rm cs}\begin{bmatrix}
        \Delta \theta \\ \Delta \phi
    \end{bmatrix},\label{eq:dsdt_hessian}
\end{align}
where the cs subscript indicates evaluating at a CS, $\Delta \theta = \theta -
\theta_{\rm cs}$, and $\Delta \phi = \phi - \phi_{\rm cs}$. The eigenvalues
$\lambda$ of Eq.~\eqref{eq:dsdt_hessian} satisfy the equation
\begin{align}
    0 &= \p{\lambda + \frac{\cos \theta_{\rm cs}}{t_{\rm s}}}\lambda - \lambda_0^2,\\
    \lambda_0^2 &\equiv \p{\alpha
        \sin \theta_{\rm cs} + g\sin I \csc^2\theta_{\rm cs}\cos \phi_{\rm cs}}
            \p{- g \sin I \cos \phi_{\rm cs}},\label{eq:def_l0_sq}\\
    \lambda &\approx -\frac{\cos \theta_{\rm cs}}{t_{\rm s}} \pm \sqrt{\lambda_0^2}.
        \label{eq:lambda_approx}
\end{align}
The stability depends only on the real part of $\lambda$ in
Eq.~\eqref{eq:lambda_approx}. $\lambda_0^2$ is a generalization of Eq.~(A4)
in, \citet{su2020} and generally has the same behavior: it is negative for CSs
1--3 and positive for CS4, as shown in Fig.~\ref{fig:lambda_full}. Thus, CS4 is
always unstable, as there will always be at least one positive solution for
$\lambda$, and the stability of CSs 1--3 are solely determined by the sign of
$\cos \theta_{\rm cs}$. From Figs.~\ref{fig:cs_locs}, we thus see that CSs 1 and
2 are stable while CS3 is unstable under the alignment torque. These
calculations justify results long used in CS literature
\citep[e.g.][]{ward1975tidal}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/2_lambdas_full.png}
    \caption{$\lambda_0^2$ (Eq.~\ref{eq:def_l0_sq}) as a function of $\eta
    \equiv \abs{g} / \alpha$ for the four CSs. The solid lines give
    $\lambda_0^2$ for the unperturbed $\phi_{\rm cs}$, and the dashed lines give
    the values for $\phi_{\rm cs}$ shifted by $60^\circ$ ($\phi_{\rm cs} =
    120^\circ$ for CS2 and $\phi_{\rm cs} = 60^\circ$ for CSs 1, 3, and 4).
    }\label{fig:lambda_full}
\end{figure}

\subsection{Spin Obliquity Evolution Driven by Alignment
Torque}\label{ss:toy_outcomes}

With the above results, we are equipped to ask questions about the dynamics of
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}): what is the long term behavior for
a general initial $\uv{s}$? If $\eta > \eta_{\rm c}$, then the only possible
final spin state is CS2, and all initial conditions will evolve asympotically
towards it. As such, we consider only $\eta < \eta_{\rm c}$, where an initial
condition can asymptotically evolve towards either CS1 or CS2. In
Fig.~\ref{fig:toy_phop}, we show which CS an initial $\theta_0$ and $\phi_0$
evolves into, where we have taken $\abs{g}t_{\rm s} = 10^{-3}$, $\eta = 0.2$,
and $I = 20^\circ$. It is clear that initial $\uv{s}$ in zone I (following the
naming in Fig.~\ref{fig:1contours}) evolve into CS1, those in zone II evolve
into CS2, while those in zone III have a probabilistic outcome. We aim to
understand each of these in turn:
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/0_eta/3stats3_5_0_2.png}
    \caption{Plot illustrating the asymptotic behavior of initial conditions for
    $\eta = 0.2$ and $I = 20^\circ$. Each dot represents an initial spin
    orientation, and the coloring of the dot indicates which Cassini State
    (legend) the system asymptotes towards.}\label{fig:toy_phop}
\end{figure}

For initial conditions in zone I, the spin circulates, and $\dot{\theta}$ is
negative everywhere during the cycle. Thus, for initial conditions in zone I,
$\theta$ decreases until the trajectory has converged to CS1. This is
intuitively reasonable, as CS1 is stable.

For initial conditions in zone II, our stability calculation in
Section~\ref{ss:tidal_equils} shows that initial conditions sufficiently near CS2
will converge to CS2. In fact, this result can be extended to all initial
conditions inside the separatrix; see Appendix~\ref{app:cs_stab2}.

For initial conditions in zone III, since there are no stable CSs in zone III,
the system must evolve through the separatrix to reach one of either CS1 or CS2.
The outcome of the separatrix encounter is effectively probabilistic and
determines the final CS\@. Intuitively, this can be understood as probabilistic
resonance capture: since $\eta \ll \eta_{\rm c}$, $\alpha \gg -g$ (the
spin-orbit precession rate, Eq.~\ref{eq:dsdt1}, and the orbit precession induced
by the perturber, Eq.~\ref{eq:dldt1}, respectively), but $\alpha \cos \theta$
can become commensurate with $-g$ if $\cos \theta \sim \eta$. This is achieved
as $\theta$ evolves from an initially retrograde obliquity through $90^\circ$
towards $0^\circ$ under the influence of the dissipative term in
Eq.~\eqref{eq:dqdt_toy}.

While similar in behavior to previous studies of probabilistic resonance capture
\citep{henrard1982, su2020}, the underlying mechanism is different: in these
previous studies, the phase space structure itself evolves and causes systems to
transition among phase space zones, while in the present scenario, a
non-Hamiltonian, dissipative perturbation causes systems to transition among
unchanging phase space zones.

\subsubsection{Analytical Calculation of Resonance Capture Probability}

The specific probabilities of the two outcomes upon separatrix encounter, a zone
III-II or a zone III-I transition, can be calculated analytically.
Figure~\ref{fig:toy_hop_manifolds} shows how the perturbative alignment torque
generates probabilistic outcomes upon separatrix encounter. We present the
interpretation of Figure~\ref{fig:toy_hop_manifolds} and the calculation of the
resonance capture probability by first giving a qualitative description of the
intuition behind the method, then presenting a calculation in good agreement with
numerical results.

We first describe the origin of the boundaries between regions of phase space
that shown in Fig.~\ref{fig:toy_hop_manifolds}. They are calculated numerically
by integrating a point infinitesimally close to CS4 forward and backward in
time. In the absence of the alignment torque, these trajectories would evolve
along the separatrix, but in the presence of the alignment torque they are
perturbed slightly and cease to overlap. It can be seen in
Fig.~\ref{fig:toy_hop_manifolds} that this splitting opens a path from zone III
into both zones I and II\@.

To understand this process quantitatively, as well as associate probabilities to
the two possible outcomes, it is important to be more quantitative. The correct
approach is to consider the evolution of the value of the \emph{unperturbed}
Hamiltonian (Eq.~\ref{eq:H}) as the spin evolves with the alignment torque. A
point in zone III evolves such that $H$ is increasing until $H \approx H_{\rm
sep}$ where $H_{\rm sep}$ is the value of $H$ along the separatrix, given by
\begin{align}
    H_{\rm sep} &\equiv H\p{\cos \theta_{\rm 4}, \phi_{\rm 4}},\nonumber\\
        &\approx -\alpha\sin I - g\cos^2 I +
            \mathcal{O}\p{\eta^2},\label{eq:def_Hsep}
\end{align}
where $\theta_4$ and $\phi_4$ are the coordinates of CS4. As the system evolves
closer to the separatrix, the change in $H$ over each circulation cycle can be
approximated by $\Delta H_-$, the change in $H$ along $\mathcal{C}_-$ (see
Fig.~\ref{fig:1contours}). In general, we can compute $\Delta H_{\pm}$ the
change over both legs of the separatrix with
\begin{equation}
    \Delta H_{\pm} \equiv \oint\limits_{\mathcal{C}_{\pm}}
        \rd{H}{t}\;\mathrm{d}t.\label{eq:def_dHpm}
\end{equation}
This can be simplified by using:
\begin{align}
    \rd{H}{t} &=
            \pd{H}{(\cos \theta)}\rd{(\cos \theta)}{t}
            + \pd{H}{\phi}\rd{\phi}{t},\nonumber\\
        &= \p{\rd{(\cos\theta)}{t}}_{\rm tide} \rd{\phi}{t},\\
    \Delta H_{\pm} &= \mp\frac{1}{t_{\rm s}}
        \int\limits_0^{2\pi} \sin^2\theta\;\mathrm{d}\phi.
\end{align}
Thus, if we evaluate $H$ every time that a trajectory crosses $\phi = 0$, we see
that it will initially be negative and increase for each circulation cycle until
the system encounters the separatrix where $H - H_{\rm sep} < \Delta H_-$.

When the system begins its separatrix-crossing orbit, $H$ at $\phi = 0$ is in
the range $\s{H_{\rm sep} - \Delta H_-, H_{\rm sep}}$; call this value $H_{\rm
i}$, as it is the initial value of the Hamiltonian on the separatrix-crossing
orbit. The two endpoints of this range are denoted in Fig.~\ref{fig:6equils} by
the black dot and cross at the left and right edges of the plot. Upon entering
this range, the trajectory encounters the separatrix, and first traverses the
vicinity of $\mathcal{C}_-$ then of $\mathcal{C}_+$, after which two possible
outcomes can occur:
\begin{itemize}
    \item If the final value of $H$ at the end of this separatrix traversal,
        denoted $H_{\rm f}$, satisfies $H_{\rm f} > H_{\rm sep}$, then the
        trajectory enters the separatrix, following the red shaded region in
        Fig.~\ref{fig:6equils}, and executes a zone III $\to$ II transition.

    \item If $H_{\rm f} < H_{\rm sep}$, then the trajectory exits the separatrix
        above CS4, following the yellow shaded region in Fig.~\ref{fig:6equils}
        and executes a zone III $\to$ I transition.
\end{itemize}
Since $H_{\rm f} = H_{\rm i} + \Delta H_+ + \Delta H_-$, we find that if $H_{\rm
i}$ is in the interval $\s{H_{\rm sep} - \Delta H_-, H_{\rm sep} - \Delta H_{-}
- \Delta H_+}$, then the system executes a III $\to$ I transition, and if it
is in the interval $\s{H_{\rm sep} - \Delta H_- - \Delta H_+, H_{\rm sep}}$,
then the system executes a III $\to$ II transition. The values of $\cos
\theta$ for which $H$ is equal to $H_{\rm sep} - \Delta H_-$, $H_{\rm sep} - \Delta
H_- - \Delta H_+$, and $H_{\rm sep}$ (CS4) for $\phi = 0$ are shown in
Fig.~\ref{fig:6equils} as the black cross, star, and dot at $\phi = 0$
respectively. Finally, if the alignment torque is weak, then $\Delta H_- \sim
\mathcal{O}\p{t_{\rm s}^{-1}}$ is small compared to the characteristic values of
$H$, and $H_{\rm i}$ can be effectively considered as uniformly distributed over
$\s{H_{\rm sep} - \Delta H_-, H_{\rm sep}}$. As such, we obtain that
\begin{equation}
    P_{\rm III \to II} = \frac{\Delta H_- + \Delta H_+}{\Delta H_-}.
        \label{eq:def_P32_toy}
\end{equation}
